\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs,hyperref}
\usepackage{pgfplots}\pgfplotsset{compat=1.18}
\usepackage[numbers,sort&compress]{natbib}
\title{A 9{,}360-Cell Pre-Registered Factorial for Frozen PLM Retrieval: Methods, Pre-Registrations, Seed-Variance Gates, and Reproducibility Infrastructure}
\author{Santiago Maniches\thanks{ORCID 0009-0005-6480-1987. TOPOLOGICA LLC.}}
\date{April 12, 2026 (v1.1)}
\begin{document}\maketitle
\begin{abstract}
We document the methodological infrastructure shared by the five papers in this compendium. A factorial of scales $\times$ panel sizes $\times$ neighbor counts $\times$ metrics $\times$ 10 seeds produces 3000-cell main results plus 3000-cell panel-shuffle null plus 3000-cell full-pool permutation null plus 180-cell cascade plus 180-cell Fisher-Rao = 9{,}360 per-cell \texttt{.npz} results, all with 10{,}000-resample percentile bootstrap CIs per stratum. Four pre-registrations with SHA256 locks (main SHA256 \texttt{139f6012\ldots}, full-pool addendum \texttt{f3864d09\ldots}, cascade, Fisher) committed before execution. A pre-committed seed-variance gate voids any (scale, $R$, $k$, metric) group where across-seed distant $F_1$ std exceeds the observed gap; all 300 main groups passed. Infrastructure is CPU-only (FAISS \cite{johnson2019billion}) and runs end-to-end in $\approx 28$ hours on a 36 GB i7 workstation. Full code release (MIT), pre-reg release (timestamped), 9{,}360 per-cell outputs released under CC-BY-4.0. This paper documents the harness rather than a specific scientific claim; it is intended as a reusable template for pre-registered PLM retrieval studies.
\end{abstract}

\section{Factorial design}
Scales: ESM-2 \cite{lin2023evolutionary} t6 (8M, 320d), t12 (35M, 480d), t30 (150M, 640d); t33 (650M, 1280d) pre-registered for GPU extension. Panel sizes $R \in \{50, 100, 250, 500, 1000\}$, class-balanced. Neighbor counts $k \in \{1, 3, 5, 10, 25\}$. Metrics: cosine (FAISS IndexFlatIP on $L_2$-normalized embeddings), Euclidean (FAISS IndexFlatL2; equivalent to cosine on $S^{d-1}$), Mahalanobis (Ledoit-Wolf \cite{ledoit2004} + eigendecomposition whitening), learned (panel-only 50-iter Adam triplet-surrogate \cite{schroff2015facenet} projection). Ten seeds 20260410-20260419.

\section{Stratification}
Per test protein, $s_{\max} = \max_{p \in \text{panel}} \langle x_{\text{test}}, x_p \rangle$ on $L_2$-normalized embeddings. Close: $s_{\max} \ge 0.95$. Moderate: $0.90 \le s_{\max} < 0.95$. Distant: $s_{\max} < 0.90$. Cuts locked in main pre-reg SHA256 \texttt{139f6012\ldots} before any evaluation. Strata with $n<100$ flagged \texttt{underpowered}.

\section{Bootstrap CIs}
10{,}000-resample percentile bootstrap per cell per stratum for $F_1$. Percentile (not BCa) chosen for computational feasibility across 9{,}360 cells; BCa jackknife is $O(n)$ per cell, percentile is $O(1)$.

\section{Seed-variance gate (pre-committed void rule)}
Per group of 10 seeds at (scale, $R$, $k$, metric): compute $\sigma^{(10)}_{\text{distant}} = \text{std}(\{F_1^{\text{distant}}_s\}_{s=1}^{10})$. If $\sigma^{(10)}_{\text{distant}} \ge \overline{F_1^{\text{close}}} - \overline{F_1^{\text{distant}}}$, void the group: the observed effect is within seed noise. All 300 main groups passed this gate. Only 2 negctrl groups and 0 fullnull groups passed (as expected; nulls should not exhibit an effect).

\begin{figure}[h]\centering
\begin{tikzpicture}\begin{axis}[width=10cm,height=5cm,xlabel={close-distant gap},ylabel={group count (of 300)},ybar,bar width=5pt,grid=major,legend pos=north west,xmin=-0.1,xmax=0.8]
\addplot[fill=red!60] coordinates {(-0.05,42)(0.00,180)(0.05,78)};
\addlegendentry{fullnull groups}
\addplot[fill=blue!60] coordinates {(0.25,8)(0.35,42)(0.45,68)(0.55,78)(0.65,62)(0.75,42)};
\addlegendentry{main groups}
\end{axis}\end{tikzpicture}
\caption{\textbf{Gap distribution across 300 groups.} Fullnull gaps concentrate at zero (red); main-factorial gaps are consistently positive and cluster at $+0.5$ to $+0.7$ (blue). The seed-variance gate voids groups where std exceeds gap; all 300 main groups pass, 0 fullnull groups pass.}\label{fig:gap_dist}
\end{figure}

\section{Pre-registrations}
\begin{enumerate}
\item \texttt{PRE\_REGISTRATION\_HOMOLOGY\_CLIFF\_v1.md} — main factorial, seed-variance gate, stratum definitions. SHA256 \texttt{139f60129d4e73dfb13444c360cc80c5b48c217d9bc87b1bf4b48b06515bcc06}.
\item \texttt{PRE\_REGISTRATION\_HOMOLOGY\_CLIFF\_ADDENDUM\_FULLNULL.md} — full-pool permutation null committed before execution after panel-only shuffle null failed its own criterion. SHA256 \texttt{f3864d097a0c611d790e6fb15a42e7efb36b2d1b103be4ec1c4f28f99d1004dc}.
\item \texttt{PRE\_REGISTRATION\_STRATIFIED\_CASCADE\_v1.md} — cascade H1.
\item \texttt{PRE\_REGISTRATION\_FISHER\_CLIFF\_v1.md} — Fisher-Rao H1.
\end{enumerate}

All harnesses verify the pre-reg file's SHA256 at runtime and abort if mismatched.

\section{Pre-registration within ML methodology literature}
Pre-registration practices \cite{nosek2018preregistration} originated in clinical trials and psychology \cite{forde2019scientific}; adoption in ML is growing but uneven \cite{guo2017calibration,ovadia2019shift}. The seed-variance gate is a contribution adapted from TOPOLOGICA prior work on GO annotation pipelines (v7 void rule); generalizing it as a domain-independent pattern for stochastic ML experiments \cite{johnson2019billion,weinberger2009distance} is one of this paper's purposes. Our approach sits alongside related methodology work \cite{schroff2015facenet,vankempen2024foldseek,urbina2022dualuse,fu2012cdhit,steinegger2017mmseqs2,carlsson2009topology,singh2007topological,edelsbrunner2010computational,amari2016,ledoit2004}.

\section{Compute and reproducibility}
Single 36 GB i7 workstation, no GPU for main factorial. Total wall time $\approx 28$ hours for 9{,}360 cells including reruns. FAISS \cite{johnson2019billion} CPU k-NN, Ledoit-Wolf via scikit-learn, PyTorch for learned projection only (CPU). Full code MIT, full data CC-BY-4.0, master SHA256 manifest at repository root.

\section{Limitations}
(L1) t33 (650M) embeddings require GPU; provided as Colab notebook.
(L2) Threshold sensitivity sweep not run (0.95/0.90 cuts fixed).
(L3) Learned metric uses a triplet-surrogate (class centroids) rather than proper triplet loss with hard negative mining; a stronger supervised baseline may exist.
(L4) Bootstrap CIs are percentile, not BCa.

\bibliographystyle{unsrtnat}\bibliography{references}\end{document}
